\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-12-12 paper/paper.tex
minor figure tweaks and to-dos

After obtaining an $N$-epoch set of observed stellar \RVs from each of the $R$ spectral orders, we combine them by modeling each \RV $v_{n,r}$ as a combination of time-dependent stellar \RV, order-dependent \RV offset, and a Gaussian noise term:
\begin{equation}
v_{n,r} = v_n + v_r + \mathcal{N}(0,\,\sigma_{n,r}^{2} + \delta_r^2 )\,,
\end{equation}
where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$.

\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$ with $v_n$ and $v_r$ analytically solved for.)}

@@

All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra. 
These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 
We obtained wavelength solutions for each spectrum from the dedicated \HARPS calibration archive, also maintained by \acronym{ESO}.

@@

If possible, the final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off an \RV drift term specified in the data header, consistent with the standard \HARPS pipeline. 
This drift is calculated from the simultaneous reference lamp and is therefore only available in certain observing modes. 
Its contribution to the \RV solution is consistently below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.

@@

We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:329968 and mbedell:331551.} 
The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra. 
Their \SNR is generally high, ranging from 100 up to 300 pix$^{-1}$ at the central wavelength regions of \HARPS.

Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the correct period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). \todo{(numbers here, address the structure in residuals)}

@@

Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:359445 and mbedell:359501.} 
\Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
In practice, this means that the 7 bluest echelle orders were consistently masked and dropped from the fit in accordance with the \SNR criteria outlined previously. 
Regardless, a large amount of data remained, allowing us to test \wobble's performance on observations with substantially different properties from the previous case.

@@

We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations. 
The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}

@@

\begin{figure}
\centering
\includegraphics[width=6.5in]{telluric_comparison}
\caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right). The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line). Residuals are shown in the lower panels. Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum. \todo{(TO DO: add y-label, fix centering)}}
\label{fig:telluric_comparison}
\end{figure}

@@

As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and \todo{xxx}.}
As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
Moreover, the composite telluric spectrum inferred from the time series of 51 Peg observations is at a much higher \SNR than the single shot telluric standard spectrum, so that low-amplitude lines which border on statistical insignificance in the standard star show up clearly in the \wobble results. 
Unlike a traditional telluric standard observation, no overhead time is required to produce these results.

@@

Another key assumption made in this work that there is no gas cell represented in the data. 
While this is true for \HARPS, other instruments like \acronym{HIRES}, \acronym{APF}, and \acronym{PFS} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \todo{(cite)}. 
Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
The gas absorption lines are fixed features at the observatory rest frame which multiply into the spectrum. 
This makes them nearly indistinguishable from the mean telluric spectrum, and they should in fact be absorbed into this component. 
If the observations in question cover a range of different airmasses, it may be necessary to explicitly add a third model component to represent the gas cell; this component would be identical to the tellurics without the airmass scaling.

@@

It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold; hence the cell. 
One simple way to deal with this would be to allow the observatory rest frame component(s) of the model to change in \RV. 
The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions.
The most extreme extension would be to include the full wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities.
\todo{(TO DO: add some citations to using tellurics for this)}

@@

Another choice we made which is almost certainly suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
In reality, every \RV should be informed by all spectral orders simultaneously. 
Implementing this would take on some extra computational cost, but being in \TF will help to mitigate this.

@@

\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

>>>>

After obtaining an $N$-epoch set of observed stellar \RVs from each of the $R$ spectral orders, we combine them by modeling each \RV $v_{n,r}$ as a combination of time-dependent stellar \RV, order-dependent \RV offset, and a Gaussian noise term:
\begin{equation}
v_{n,r} = v_n + v_r + \mathcal{N}(0,\,\sigma_{n,r}^{2} + \delta_r^2 )\,,
\end{equation}
where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$. 
The order-dependent \RV offset should be small (and indeed generally is consistent with zero in the solutions), but is included in the model to capture any color-dependent effects such as non-negligible atmospheric variability during the exposure \citep[see e.g.][]{Blackwell2017}.

@@

All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra.\footnote{Based on observations made with ESO Telescopes at the La Silla Paranal Observatory under programme IDs 091.C-0271, 183.C-0437, 191.C-0505, 072.C-0488, 089.C-0497, 099.C-0880, and 60.A-9700.} 
These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 
We obtained wavelength solutions for each spectrum from the dedicated \HARPS calibration archive, also maintained by \acronym{ESO}.

@@

If possible, the final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off an \RV drift term specified in the data header, consistent with the standard \HARPS pipeline. 
This drift is calculated from the simultaneous reference lamp and is therefore only available in certain observing modes. 
Its contribution to the \RV solution is usually below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.

@@

We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.
The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra. 
Their \SNR is generally high, ranging from 100 up to 300 pix$^{-1}$ at the central wavelength regions of \HARPS.

Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). \todo{(numbers here, address the structure in residuals)}

@@

Next, we tested \wobble with 306 epochs of \Mdwarf spectra.
\Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
In practice, this means that the 7 bluest echelle orders were consistently masked and dropped from the fit in accordance with the \SNR criteria outlined previously. 
Regardless, a large amount of data remained, allowing us to test \wobble's performance on observations with substantially different properties from the previous case.

@@

We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations. 
The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.

@@

\begin{figure}
\centering
\includegraphics[width=6.5in]{telluric_comparison}
\caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right). The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line). Residuals to the fit are shown in the lower panels. Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
\label{fig:telluric_comparison}
\end{figure}

@@

As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star HR 3090 taken by \HARPS at an \SNR of 130. 
As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
Moreover, the composite telluric spectrum inferred from the time series of 51 Peg observations is at a much higher \SNR than the single shot telluric standard spectrum, so that low-amplitude lines which border on statistical insignificance in the standard star show up clearly in the \wobble results. 
Unlike a traditional telluric standard observation, no overhead time is required to produce these results.

@@

Another key assumption made in this work that there is no gas cell represented in the data. 
While this is true for \HARPS, other instruments such as \acronym{HIRES}, \acronym{PFS}, and \acronym{APF} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \citep{Butler1996, Crane2010, Vogt2014}. 
Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
The gas absorption lines are fixed features at the observatory rest frame which multiply into the spectrum. 
This makes them nearly indistinguishable from the mean telluric spectrum, and they should in fact be absorbed into this component. 
If the observations in question cover a range of different airmasses, it may be necessary to explicitly add a third model component to represent the gas cell; this component would be identical to the tellurics without the airmass scaling.

@@

It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold; hence the cell. 
One simple way to deal with this would be to allow the observatory rest frame component(s) of the model to change in \RV. 
The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions.
The most extreme extension would be to include corrections to the wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities. 
This would rely on the use of telluric features as fixed calibration sources \citep{Seifahrt2010}.

@@

Another choice we made which is likely suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
In reality, every stellar \RV should be informed by all spectral orders simultaneously. 
Implementing this would take on some extra computational cost, but being in \TF will help to mitigate this. 
However, the reality of color-dependent atmospheric dispersion effects means that in practice each order may have a slightly different effective \RV, since the flux-weighted average observation times will be different \citep{Blackman2017}. 
Thus for some situations allowing the orders to have independent \RV estimates may be the best choice.

@@

\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.
